Ab Initio Effective One-Electron Potential Operators: Applications for Charge-Transfer Energy in Effective Fragment Potentials

### 2019-11-15 papers/2_oep_technique/paper-r2.tex
CT paper revision added new sections

Four selected molecular pairs composed of H$_2$O, CH$_3$OH, NH$_4^+$ and NO$_3^-$
(Figure 1) were chosen to validate the OEP\hyp{}based model of the CT energy
at the PT/HF formulation level. Structures of the molecular dimers were first 
energy\hyp{}optimized at the HF/6-31+G(d,p) level by using
the Gaussian 16 quantum chemistry program package.\cite{Gaussian16}
To obtain the PT/HF benchmark CT energy, it was estimated
from the following approximation:
%
\begin{equation}\label{e:ct-pthf-ref}
 \Delta E^{\rm CT}_{\rm PT/HF} \cong \Delta E^{\rm Pol}_{\rm DDS/HF} 
 + \frac{1}{2} {\bf F}^{\rm T} \cdot {\bf D}^{-1} \cdot {\bf F} \;,
\end{equation}
%
where $\Delta E^{\rm Pol}_{\rm DDS/HF}$ is the Hartree\hyp{}Fock polarization energy
defined
according to the density decomposition scheme (here referred as to the DDS) 
proposed by Mandado and Hermida\hyp{}Ram{\'o}n,\cite{Mandado.Hermida-Ramon.JCTC.2011} 
whereas the remainder term is the 
approximate pure induction energy due to the polarizability effects.\cite{Li.Netzloff.Gordon.JCP.2006}
The DDS polarization energy encompasses all the charge delocalization
effects that are not associated with the Pauli exclusion principle
in the first order, and is computed by
%
\begin{multline}\label{e:de-pol-dds}
\Delta E^{\rm Pol}_{\rm DDS/HF} \equiv \Delta E^{\rm HF} \\
 - \left(
   \Delta E^{\rm Coul}_{\rm DDS/HF}
 + \Delta E^{\rm Exch}_{\rm DDS/HF}
 + \Delta E^{\rm Rep}_{\rm DDS/HF} 
   \right)\;.
\end{multline}
%
In the above, $\Delta E^{\rm Coul}_{\rm DDS/HF}$ is the pure Coulombic interaction energy
between the unperturbed (non\hyp{}interacting) charge densities of the monomers in the complex,
and
$\Delta E^{\rm Exch}_{\rm DDS/HF}$ and $\Delta E^{\rm Rep}_{\rm DDS/HF}$
are the exchange and Pauli\hyp{}repulsion interaction energies, respectively. 
At the HF level of theory, their sum is in fact
equivalent to the first\hyp{}order exchange\hyp{}repulsion energy 
in the intermolecular perturbation theory with exchange of Hayes and Stone.\cite{Hayes.Stone.MolPhys.1984} 
It was also shown by 
Mandado and Hermida\hyp{}Ram{\'o}n,\cite{Mandado.Hermida-Ramon.JCTC.2011}
that the Coulombic and exchange\hyp{}repulsion term
in the DDS method reproduce the first\hyp{}order Coulombic and exchange\hyp{}repulsion
interaction energy in the symmetry adapted perturbation theory\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994}
(SAPT)
with accuracy below 0.5 kcal/mole.
Therefore, Eq.~\eqref{e:ct-pthf-ref} is believed to be sufficient
in order to compute the PT/HF CT energy, because there is no dispersion
contribution to the DDS polarization energy at the HF level.

The pure induction energy was estimated by using the distributed
dipole\hyp{}dipole polarizability approximation by using the computational
methodology by Li et al.,\cite{Li.Netzloff.Gordon.JCP.2006} 
in which ${\bf F}$ constitutes the electric fields aligned in a super\hyp{}vector 
according to each polarizable center, i.e. ${\bf F} = \{ {\bf F}_1, {\bf F}_2, \ldots\}$,
whereas
the ${\bf D}$ matrix is defined as
%
\begin{equation}
 {\bf D}_{ab} = 
\left\{\begin{matrix}
{\BM\alpha}_a^{-1} &\text{if $a=b$} \\ 
0                  &\text{if $a\ne b \land a,b \in$ same molecule}\\ 
-{\bf T}_{ab}      &\text{if $a\ne b \land a,b \notin$ same molecule}
\end{matrix}\right.
\end{equation}
%
and ${\bf T}_{ab}$ is the dipole\hyp{}dipole interaction tensor
given by
%
\begin{equation}
 {\bf T}_{ab} \equiv \frac{3}{r_{ab}^5} {\bf r}_{ab} \otimes {\bf r}_{ab}
  - \frac{1}{r_{ab}^3} {\bf 1} \;,
\end{equation}
%
with ${\bf r}_{ab} \equiv {\bf r}_{a} - {\bf r}_{b}$ being the
relative distance between the LMO centroids.
The distributed polarizabilities {\bf\textalpha} ${\boldsymbol{\mathrm{\upalpha}}}_a$ centered at the LMO centroids
were computed by utilizing the coupled\hyp{}perturbed Hartree\hyp{}Fock
method.\cite{McWeeny.RevModPhys.1960,Dodds.McWeeney.Sadlej.MolPhys.1977} 
Orbital localization was performed by using the Boys method.\cite{Boys.RevModPhys.1960}

>>>>

Four selected molecular pairs composed of H$_2$O, CH$_3$OH, NH$_4^+$ and NO$_3^-$
(Figure 1) molecules or ions were chosen to validate the OEP\hyp{}based model of the CT energy
at the PT/HF formulation level. Structures of the molecular dimers were first 
energy\hyp{}optimized at the HF/6-31+G(d,p) level by using
the Gaussian 16 quantum chemistry program package.\cite{Gaussian16}
To obtain the PT/HF benchmark CT energy, it was estimated
from the following approximation:
%
\begin{equation}\label{e:ct-pthf-ref}
 \Delta E^{\rm CT}_{\rm PT/HF} \cong \Delta E^{\rm Pol}_{\rm DDS/HF} 
 - \Delta E^{\rm Ind} \;,
\end{equation}
%
where $\Delta E^{\rm Pol}_{\rm DDS/HF}$ is the Hartree\hyp{}Fock polarization energy
defined
according to the density decomposition scheme (here referred as to the DDS) 
proposed by Mandado and Hermida\hyp{}Ram{\'o}n,\cite{Mandado.Hermida-Ramon.JCTC.2011} 
whereas $\Delta E^{\rm Ind}$ term is the 
approximate pure induction energy due to the polarizability effects.\cite{Stone.TheTheoryOfIntermolecularForces.1996}
The DDS polarization energy encompasses all the charge delocalization
effects that are not associated with the Pauli exclusion principle
in the first order, and is computed by
%
\begin{multline}\label{e:de-pol-dds}
\Delta E^{\rm Pol}_{\rm DDS/HF} \equiv \Delta E^{\rm HF} \\
 - \left(
   \Delta E^{\rm Coul}_{\rm DDS/HF}
 + \Delta E^{\rm Exch}_{\rm DDS/HF}
 + \Delta E^{\rm Rep}_{\rm DDS/HF} 
   \right)\;.
\end{multline}
%
In the above, $\Delta E^{\rm Coul}_{\rm DDS/HF}$ is the pure Coulombic interaction energy
between the unperturbed (non\hyp{}interacting) charge densities of the monomers in the complex,
and
$\Delta E^{\rm Exch}_{\rm DDS/HF}$ and $\Delta E^{\rm Rep}_{\rm DDS/HF}$
are the exchange and the Pauli\hyp{}repulsion interaction energies, respectively. 
At the HF level of theory, the sum of the latter two terms is in fact
equivalent to the first\hyp{}order exchange\hyp{}repulsion energy 
in the intermolecular perturbation theory with exchange of Hayes and Stone.\cite{Hayes.Stone.MolPhys.1984} 
It was also shown by 
Mandado and Hermida\hyp{}Ram{\'o}n,\cite{Mandado.Hermida-Ramon.JCTC.2011}
that the Coulombic and exchange\hyp{}repulsion term
in the DDS method reproduce the first\hyp{}order Coulombic and exchange\hyp{}repulsion
interaction energy in the symmetry adapted perturbation theory\cite{Jeziorski.Moszynski.Szalewicz.ChemRev.1994}
(SAPT)
with accuracy below 0.3 kcal/mole.
Therefore, Eq.~\eqref{e:ct-pthf-ref} is believed to be sufficient
in order to compute the PT/HF CT energy, because there is no dispersion
contribution to the DDS polarization energy at the HF level.

The pure induction energy was estimated by using the distributed
dipole\hyp{}dipole polarizability approximation according to the computational
methodology of Li et al.,\cite{Li.Netzloff.Gordon.JCP.2006}
in which
%
\begin{equation}\label{e:ind-ref}
 \Delta E^{\rm Ind} \approx - \frac{1}{2} {\bf F}^{\rm T} \cdot {\bf D}^{-1} \cdot {\bf F} \;.
\end{equation}
%
In the above equation, ${\bf F} = \{ \ldots, {\bf F}({\bf r}_a), \ldots\}$
is a super\hyp{}vector of electric fields at the $a$th distributed site
due to the surrounding unperturbed molecular charge densities in the complex
(i.e., the charge\hyp{}density of the molecule that contains $a$th center
is not included in ${\bf F}({\bf r}_a)$).
%The electric field components are
%aligned in a super\hyp{}vector 
%according to each polarizable center, i.e. ${\bf F} = \{ \ldots, {\bf F}({\bf r}_a), \ldots\}$,
%where $a$ labels the LMO centroids.
The ${\bf D}$ matrix is defined as
%
\begin{equation}
 {\bf D}_{ab} = 
\left\{\begin{matrix}
{\BM\upalpha}_a^{-1} &\text{if $a=b$} \\ 
0                    &\text{if $a\ne b \land a,b \in$ same molecule}\\ 
-{\bf T}_{ab}        &\text{if $a\ne b \land a,b \notin$ same molecule}
\end{matrix}\right.
\end{equation}
%
and ${\bf T}_{ab}$ is the dipole\hyp{}dipole interaction tensor
given by
%
\begin{equation}
 {\bf T}_{ab} \equiv \frac{3}{r_{ab}^5} {\bf r}_{ab} \otimes {\bf r}_{ab}
  - \frac{1}{r_{ab}^3} {\bf 1} \;,
\end{equation}
%
with ${\bf r}_{ab} \equiv {\bf r}_{a} - {\bf r}_{b}$ being the
relative distance between the LMO centroids.
The electric fields at the LMO centroids were computed from
the unperturbed HF charge\hyp{}density distributions
%
\begin{equation}
 {\bf F}({\bf r}_a) = \sum_x^{\rm At} \frac{Z_x}{r_{ax}^3} {\bf r}_{ax} 
  + 2 \sum_{\alpha\gamma} D_{\alpha\gamma} 
%    \tBraKet{\gamma}{\frac{{\bf r}_a - {\bf r}}{\vert {\bf r}_a - {\bf r} \vert^3} }{\alpha}_{\bf r} \;.
  \left< \gamma \left| 
    \frac{{\bf r}_a - {\bf r}}{\vert {\bf r}_a - {\bf r} \vert^3}
               \right| \alpha \right>_{\bf r} \;.
\end{equation}
%
The distributed anisotropic dipole\hyp{}dipole
polarizabilities ${\BM{\upalpha}}_a$ centered at the LMO centroids
were computed by utilizing the coupled\hyp{}perturbed Hartree\hyp{}Fock
method.\cite{McWeeny.RevModPhys.1960,Dodds.McWeeney.Sadlej.MolPhys.1977} 
Orbital localization was performed by using the Boys method.\cite{Boys.RevModPhys.1960}
